---
title: "Using LRCDE"
author: "Edmund R Glass"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using LRCDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

LRCDE is a linear regression deconvolution package.  It is designed to do the following:

* Cell type-specific differential expression estimation
* output the observed power of each cell type-specific difference estimate
    * Very small observed differences _might_ be real... yet underpowered.
* Allow simulation of cell proportions matrices with:
    * desired level of variability (targeted standard deviation) across target cell 1
    * desired level of condition number across entire cell proportions matrix (for single study group)
* Allow quick easy simulation of cell type-specific expression matrices
* Allow simulation of heterogeneous observations with targeted level residual variability (MSE target)

The following steps allow for testing of the simulation facilities and the lrcde function.

## Initial setup

You must start by declaring the working directory where a .CSV output file will be written.

Then you declare a few parameter values for the simulation functions to use.

```{r, results="hide"}
library(lrcde) # Load the lrcde package

# setwd("/home/your.user.name/output.directory") # change this to suit your own setup and preferences

# Custom parameters to model:

    n.samps = 15 # Sample size per group
    # Mean Squared Error to model (actual average MSE per gene will be a small fraction of this)
     mse2model.vec  = c( 0.2)
     
# Cell proportion parameterss to model:
    # Target cell standard deviation (across samples) to model:
    cell.sd.2.model = c( 0.08 )
    # Condition number (kappa) to target (resulting kappa will be approximate):
    kappp.2.model  = c( 71500 )
    # Number of cells to simulate:
    n.cells  = c( 5 )
    # The "target" cell (the one with the fold change) for simulations
    cell.p = 1

# Cell expression params to model:

    # Base level cell expressions to model:
    base.expr.vec    = c( 2  )
    # Cell level differential expression to model:
    diff.2.model.vec = c( 0.6, 0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.3, 1.4, 1.5 )

```
As you can see, there will be 10 different effect-sizes models.  Since only a single mean squared error (MSE) is being modeled, then there will be a simulated heterogeneous matrix with only 10 features modeled.


## Data Simulation

Once paramters have been declared, you may simulate data.

```{r, results="hide"}
###############################################################################
# Simulate data:
    
# For the sake of reproducibility, you should set a seed:
seed2set = (11221963)

# Group indicator vector (group membership indicator - should NOT be a factor here!):
groups = c(rep(1,n.samps), rep(2,n.samps) )

# Now simulate cell proportions (note use of set.seed):
set.seed(seed2set)
cell.props.1 = cell.props.target( n.cells             # Number of cells to simulate
                                , n.samps           # Number of samples (per group) to simulate
                                , cell.sd.2.model   # Standard deviation of target cell
                                , kappp.2.model )   # Condition number of cell.props (per group)

# How did the cell proportions turned out ?:
apply(cell.props.1, 2, sd)                          # Standard deviation of target cell (cell.p)
kappa(t(cell.props.1)%*%cell.props.1, exact=TRUE)   # actual kappa (condition number)

# Stack control and cases (identical) cell proportions:    
cell.props = rbind( cell.props.1, cell.props.1 )

# Sim cell level expression (gold-standard).
cell.expr = custom.sim.cell.expr(   n.cells           # Number of cell types being simulated
                                  , base.expr.vec     # The "base" expression level to model
                                  , diff.2.model.vec  # Differential expression to model
                                  , cell.p            # Target cell to modify in cases
                                  , length(mse2model.vec) )

# Sim residuals:
set.seed(seed2set)
resids = custom.resids.synthetic(    mse2model.vec    # Actual MSE will be small fraction of this
                                   , groups           # groups vector
                                   , diff.2.model.vec # Included to get matrix size correct
                                   , base.expr.vec    # Included to get matrix size correct
                                   , adjuster=1 )     # Scaling factor for MSE target

# Sim het obs:
het.obs = het.from.synthetic(   cell.props     # The entire cell proportions matrix
                              , cell.expr      # Cell type-specific expressions matrix
                              , resids         # Simulated residuals matrix
                              , groups )       # groups membership vector

colnames( het.obs ) = 1:dim( het.obs )[2]         # LRCDE expects to see feature names
colnames( cell.props ) = 1:dim( cell.props )[2]   # LRCDE expects to see cell type names
```

Congratualtions!  You just simulated some data.


## Running the LRCDE function

Using the data that you just simulated, you now call the lrcde function.  The main function in the LRCDE package.

The following three parameter will automatically default to the following values if undeclared in the call to lrcde.
The values shown below are the recommended settings.

DO NOT standardize or median center differential expression estiamtes when using LRCDE... In fact, I should remove these features from the code.

```{r, results="hide"}
    ###############################################################################
    # Use these for LRCDE since power calculation is meaningless if
    #      differences are transformed but standard errors are not:
    stdz=FALSE; medCntr=FALSE; nonNeg=TRUE
    ###############################################################################
```

The 'method to use' will default to 'dual' since this is the only method implemented.

It is good to specically declare the name of your output file (leave the .csv extension).

The 'alternative' can be one of 'two.sided', 'greater', or 'less'.  Default is 'two.sided'.

```{r, results="hide"}
    method2use="dual" # Which type of deconvolution to run (dual is only thing implemented)

    lrcde.output.file   = paste0( "lrcde_sim_example.csv"  )
    alternative='two.sided' # One of "two.sided", "greater", or "less"

    # Run LRCDE:
    return.list = lrcde(  het.obs, cell.props, groups
                          , output.file = lrcde.output.file
                          , medCntr     = medCntr
                          , stdz        = stdz
                          , nonNeg      = nonNeg
                          , method      = method2use
                          , direction   = alternative
    )
```

The lrcde function will output each cell type name after it has finished analyzing all of the features across that cell type.

The 'return.list' is a 2 item list.  Item 1 is a dataframe of results of the power analysis.  Item 2 is another list of the parameters and their values used in the call to lrcde

Take a look at the output data frame for the 10 feature simulation that you just ran

```{r, results="hide"}
    result.frame = return.list[[1]]
    result.frame
```

Notice the 'power' column.  This has the observed power for the observed cell type-specific difference estimates.

Only the target cell (cell 1 here) will have significant power numbers.

The features with the larget differences will have higher power (as it should be) since each feature has approximately the same MSE (residual sizes).

Other vignettes include fancier examples.

